Short-term neuronal and synaptic plasticity act in synergy for deviance detection in spiking networks

### 2023-01-06 paper-1/manuscript.tex
Moved figures to/near the start of their text
Added SI figures
+todo: Discussion
Bugfixes
Rev: remove RSI, MMI

\begin{figure*}%
    \centering
    \includegraphics{fig2.pdf}
    \caption{}
    \label{fig2}
\end{figure*}
\textbf{Fig. 2} Deviance detection in our model. \textbf{a} Sample network target response intensity by trial type. Each horizontal line of the plot summarizes the network response in one target trial in terms of spikes per integration time step. Target trials are taken from the std (400), msc (100), and dev (100) sequences, and sorted by sequence time (earliest trials are at the bottom of each section). \textbf{b} Response summary, showing average network-wide spike counts in target trials in each sequence. The underlying data are the same as in panel a. Note how the dev response is very large, followed by an intermediate msc response and a low std response. \textbf{c} Mismatch, repetition suppression, and deviance detection indices (see Equation \ref{eqn-ddi}) across networks and stimuli (n = 120). In this and later boxplots, the median is indicated with an orange line, notches indicate the 95\% confidence interval for the median calculated by bootstrap with 10000 iterations, boxes indicate the inter-quartile range (IQR), whiskers extend to 1.5 times the IQR or to the data extrema, whichever is less, and fliers indicate individual data points beyond the whisker end points

We define deviance detection as a network's ability to pick out unexpected inputs from an otherwise homogeneous or otherwise unsurprising stream of inputs. Following related research \citep{Kubota2021-dx,Harms2014-ah,Jacobsen2001-sc}, we were careful to exclude effects due to stimulus identity (i.e., avoiding direct comparisons between different stimuli) or due to adaptation to frequent stimulus presentation (i.e., avoiding direct comparisons of A in the dev sequence to A in the std sequence). In other words, we used the msc sequence as a neutral baseline where the target stimulus was presented no more often than in the dev sequence, but no expectation of a competing stimulus B could be formed. To quantify deviance detection, we compared the mean number of spikes $R^A$ fired across the network in response to target stimulus A in the dev and msc sequences, defining a deviance detection index (DDI) as
\begin{equation}
    DDI = \frac{R^A_{dev} - R^A_{msc}}{R^A_{dev} + R^A_{msc}} \label{eqn-ddi}
\end{equation}
We define related indices, contrasting the responses in msc against std, and in dev against std, analogously. Since contrasts of both kinds have been called "stimulus-specific adaptation" or SSA in the literature \citep{Kubota2021-dx,Taaseh2011-gg}, we chose to avoid such ambiguity, using instead the term "repetition suppression" (RSI, $R^A_{msc} - R^A_{std}$) to refer to the reduction in response induced mainly by a difference in presentation frequency while controlling for context, and the term "mismatch" (MMI, $R^A_{dev} - R^A_{std}$) to refer to the uncontrolled contrast, which captures a combination of repetition suppression and deviance detection.

@@

We could readily identify similar patterns emerging across most networks and stimulus assignments. To show that deviance detection is robust to network and stimulus identity, we calculated the DDI and related indices mentioned above for each network, and show them in a summary plot in Figure \ref{fig2}c. We found that our model reliably produced positive mismatch indices (T = 6.67e+03, p = 9.26e-16, median 0.253; one-sided Wilcoxon signed-rank test, n = 120) while both RSI (T = 6.19e+03, p = 9.76e-12, median 0.135) and DDI (T = 5.97e+03, p = 4.75e-10, median 0.114) were also positive, despite some outliers. This shows that our model was well behaved and suitable for an investigation of deviance detection.

\begin{figure*}%
    \centering
    \includegraphics{fig3.pdf}
    \caption{}
    \label{fig3}
\end{figure*}
\textbf{Fig. 3} Deviance detection under model ablation. \textbf{a} Sample network responses to target stimulation, with the deviance detection index in each instance noted in the plot titles. Lines show the average (trial mean +- SEM, left axis) network-wide spike counts in target trials in each sequence. Boxes summarize the spike count totals per trial (right axis), the means of which form the basis for index calculations (see panel c). Note the different axis scaling on the top row (without TA) and the bottom row (with TA). \textbf{b} Repetition suppression indices $<$msc - std$>$ across networks and stimuli. Asterisks indicate data greater than 0, established with a Wilcoxon signed-rank test at a significance level 0.05. See main text for detailed statistics. \textbf{c} Deviance detection indices $<$dev - msc$>$ across networks and stimuli

Having confirmed the presence of deviance detection in our model, we turned to an ablation approach to try to identify how the two short-term plasticity mechanisms in the model, STD and TA, enabled the networks to learn relevant regularities. In the remainder of the paper, we will proceed to detail the processes leading to deviance detection on a sample network, then confirm the highlighted trends statistically with reference to the full set of 120 networks and stimulus assignments. The sample network was chosen to be roughly representative of the population as follows: We assessed each network's average response to stimulation with A and B in the msc sequence and calculated z-scores with respect to the population. Considering only networks whose absolute z-score was less than 1 for both A and B responses, we hand-picked a sample network and stimulus assignment that exhibited both an increased DDI in STD+TA over TA only, and positive RSI in both STD+TA and TA only conditions.

@@

The RSI (Figure \ref{fig3}b) was exactly 0 as expected in the ablation with no plasticity, and positive with any plasticity applied (STD only: T = 4.97e+03, p = 0.000222, median 0.0039; TA only: T = 5.99e+03, p = 3.3e-10, median 0.097; STD+TA: T = 6.19e+03, p = 9.76e-12, median 0.135; one-sided Wilcoxon signed-rank tests, n = 120). Since the RSI measures the degree of adaptation to frequent stimulus presentation, this indicates that both plasticity mechanisms were capable of imbuing the networks with some level of history dependence.

Finally, across networks and stimuli, the DDI (Figure \ref{fig3}c) was exactly 0 with no plasticity, not significantly different from 0 with STD only (T = 3.2e+03, p = 0.262, median -2.1e-04; two-sided Wilcoxon, n = 120), but clearly positive for TA only (T = 5.85e+03, p = 3.1e-09, median 0.077; one-sided Wilcoxon, n = 120) and the full model (T = 5.97e+03, p = 4.75e-10, median 0.114). To our surprise, the DDI of the full model was significantly greater than that of either plasticity mechanism alone (STD: T = 6.1e+03, p = 4.95e-11; TA: T = 4.87e+03, p = 0.000604), and greater too than the sum of the DDI values across both STD only and TA only models (T = 5.12e+03, p = 4.52e-05). This suggests that TA and STD act not as independent mechanisms in this paradigm, but rather interact constructively to enhance the network's ability to encode input regularities.

@@

@@

\begin{figure*}%
    \centering
    \includegraphics{fig7.pdf}
    \caption{}
    \label{fig7}
\end{figure*}
\textbf{Fig. 7} Adding STD to the model maintains or slighly enhances the effects of TA.
\textbf{a} Median contrast (dev - msc) in the non-target response (left) and in $V_{TA}$ at the start of B trials (right) among early and late neurons, respectively, across networks and stimuli, in the full model with both STD and TA. Compare to Figure \ref{fig6}c-d.
\textbf{b} Median contrast (dev - msc) in the non-target response magnitude across networks and stimuli in the full model (horizontal axis) and the ablated model with TA only (vertical axis). The red line is the identity function; points above this line represent networks with greater response reduction (dev - msc) in the full model than in the ablated model. The subtraction of the two quantities in the box plot below shows that the non-target response tends to be reduced more in the full model.
\textbf{c} Non-target response contrast, plotted against its Pearson correlation coefficient with the corresponding contrast in $V_{TA}$ at the start of target trials, in the full model. Compare to Figure \ref{Fig5}d.
\textbf{d} Median contrast (dev - msc) in $V_{TA}$ at the start of target trials across networks and stimuli in the full and ablated models, arranged as in panel b. The subtraction of the two quantities in the box plot below shows that $V_{TA}$ tends to be reduced more in the full model.
\textbf{e} $\Delta V_{TA}$ at the start of target trials, plotted against its Pearson correlation coefficient with the corresponding contrast in the target response magnitude, in the full model. Compare to Figure \ref{Fig4}e.
\textbf{f} Median contrast (dev - msc) in the target response magnitude across networks and stimuli in the full and ablated models, arranged as in panel b. The subtraction of the two quantities in the box plot below shows that the target response tends to be increased more in the full model.

@@

We had good reasons for this assumption: Firstly, referring back to Figure \ref{fig3}, notice that the ablated model with STD only produced a positive RSI (i.e., the target response in msc was greater than in std), but not a significant DDI (i.e., no difference between target dev and msc responses). Secondly, we note that updates to the spike-triggered depression variable $x_j$ (cf. Equation \ref{eqn-xsyn}) would decay to a negligible $e^\frac{-1 s}{\tau_x} = 0.0013$ times the original value within two trials, making it unlikely that inputs separated by intervening trials could be "remembered" in the network. Conversely, the frequently presented stimulus B may have been able to lead to a build-up of depression strong enough to suppress its own response.

@@

@@

% \begin{appendices}

% \section{Section title of first appendix}\label{secA1}

@@

% \end{appendices}

>>>>

We define deviance detection as a network's ability to pick out unexpected inputs from an otherwise homogeneous or otherwise unsurprising stream of inputs. Following related research \citep{Kubota2021-dx,Harms2014-ah,Jacobsen2001-sc}, we were careful to exclude effects due to stimulus identity (i.e., avoiding direct comparisons between different stimuli) or due to adaptation to frequent stimulus presentation (i.e., avoiding direct comparisons of A in the dev sequence to A in the std sequence). In other words, we used the msc sequence as a neutral baseline where the target stimulus was presented no more often than in the dev sequence, but no expectation of a competing stimulus B could be formed. To quantify deviance detection, we compared the mean number of spikes fired per neuron in response to target stimulus A in the dev and msc sequences, denoted as $R^A_{dev}$ and $R^A_{msc}$, defining a deviance detection index (DDI) as
\begin{equation}
    DDI = \frac{R^A_{dev} - R^A_{msc}}{R^A_{dev} + R^A_{msc}} \label{eqn-ddi}
\end{equation}

\begin{figure*}%
    \centering
    \includegraphics{fig2.pdf}
    \caption{}
    \label{fig2}
\end{figure*}
\textbf{Fig. 2} Deviance detection in our model. \textbf{a} Sample network target response intensity by trial type. Each horizontal line of the plot summarizes the network response in one target trial in terms of spikes per integration time step. Target trials are taken from the std (400), msc (100), and dev (100) sequences, and sorted by sequence time (earliest trials are at the bottom of each section). \textbf{b} Response summary, showing average network-wide spike counts in target trials in each sequence. The underlying data are the same as in panel a. Note how the dev response is very large, followed by an intermediate msc response and a low std response.
\textbf{c} Response sizes $R^A_{dev}$ and $R^A_{std}$, normalized by $R^A_{msc}$, across networks and stimuli (n = 120). In this and later boxplots, the median is indicated with an orange line, notches indicate the 95\% confidence interval for the median calculated by bootstrap with 10000 iterations, boxes indicate the inter-quartile range (IQR), whiskers extend to 1.5 times the IQR or to the data extrema, whichever is less, and fliers indicate individual data points beyond the whisker end points.
\textbf{d} Deviance detection indices (DDI, see Equation {eqn-ddi}) across networks and stimuli

@@

We could readily identify similar patterns emerging across most networks and stimulus assignments. To show that deviance detection is robust to network and stimulus identity, we calculated the average spike count per neuron per target trial for each network and stimulus assignment, and compared these across conditions. Figure \ref{fig2}c shows a summary of the average dev and std response sizes, normalized by the response in the msc sequence. The data clearly confirmed that dev responses were larger than msc (T = 6.38e+03, p = 2.75e-13; one-sided Wilcoxon signed-rank test, n = 120), and std responses smaller (T = 687, p = 6.42e-15). Accordingly, the DDI (Figure \ref{fig2}d) was positive in most networks (T = 5.97e+03, p = 4.75e-10, median 0.114). This shows that our model was well behaved and suitable for an investigation of deviance detection.

Having confirmed the presence of deviance detection in our model, we turned to an ablation approach to try to identify how the two short-term plasticity mechanisms in the model, STD and TA, enabled the networks to learn relevant regularities. In the remainder of the paper, we will proceed to detail the processes leading to deviance detection on a sample network, then confirm the highlighted trends statistically with reference to the full set of 120 networks and stimulus assignments. The sample network was chosen to be roughly representative of the population as follows: We assessed each network's average response to stimulation with A and B in the msc sequence and calculated z-scores with respect to the population. Considering only networks whose absolute z-score was less than 1 for both A and B responses, we hand-picked a sample network and stimulus assignment that exhibited both an increased DDI in STD+TA over TA only, and greater msc than std responses in both STD+TA and TA only conditions.

\begin{figure*}%
    \centering
    \includegraphics{fig3.pdf}
    \caption{}
    \label{fig3}
\end{figure*}
\textbf{Fig. 3} Deviance detection under model ablation. \textbf{a} Sample network responses to target stimulation, with the deviance detection index in each instance noted in the plot titles. Lines show the average (trial mean +- SEM, left axis) network-wide spike counts in target trials in each sequence. Boxes summarize the spike count totals per trial (right axis), the means of which form the basis for index calculations (see panel c). Note the different axis scaling on the top row (without TA) and the bottom row (with TA). \textbf{b} Deviance detection indices across networks and stimuli. Asterisks indicate data greater than 0, established with a Wilcoxon signed-rank test at a significance level 0.05. See main text for detailed statistics

@@

Finally, across networks and stimuli, the DDI (Figure \ref{fig3}c) was exactly 0 as expected in the ablation with no plasticity, not significantly different from 0 with STD only (T = 3.2e+03, p = 0.262, median -2.1e-04; two-sided Wilcoxon, n = 120), but clearly positive for TA only (T = 5.85e+03, p = 3.1e-09, median 0.077; one-sided Wilcoxon, n = 120) and the full model (T = 5.97e+03, p = 4.75e-10, median 0.114). To our surprise, the DDI of the full model was significantly greater than that of either plasticity mechanism alone (STD: T = 6.1e+03, p = 4.95e-11; TA: T = 4.87e+03, p = 0.000604), and greater too than the sum of the DDI values across both STD only and TA only models (T = 5.12e+03, p = 4.52e-05). This suggests that TA and STD act not as independent mechanisms in this paradigm, but rather interact constructively to enhance the network's ability to encode input regularities.

@@

\begin{figure*}%
    \centering
    \includegraphics{fig7.pdf}
    \caption{}
    \label{fig7}
\end{figure*}
\textbf{Fig. 7} Adding STD to the model maintains or slighly enhances the effects of TA.
\textbf{a} Median contrast (dev - msc) in the non-target response (left) and in $V_{TA}$ at the start of B trials (right) among early and late neurons, respectively, across networks and stimuli, in the full model with both STD and TA. Compare to Figure \ref{fig6}c-d.
\textbf{b} Median contrast (dev - msc) in the non-target response magnitude across networks and stimuli in the full model (horizontal axis) and the ablated model with TA only (vertical axis). The red line is the identity function; points above this line represent networks with greater response reduction (dev - msc) in the full model than in the ablated model. The subtraction of the two quantities in the box plot below shows that the non-target response tends to be reduced more in the full model.
\textbf{c} Non-target response contrast, plotted against its Pearson correlation coefficient with the corresponding contrast in $V_{TA}$ at the start of target trials, in the full model. Compare to Figure \ref{fig5}d.
\textbf{d} Median contrast (dev - msc) in $V_{TA}$ at the start of target trials across networks and stimuli in the full and ablated models, arranged as in panel b. The subtraction of the two quantities in the box plot below shows that $V_{TA}$ tends to be reduced more in the full model.
\textbf{e} $\Delta V_{TA}$ at the start of target trials, plotted against its Pearson correlation coefficient with the corresponding contrast in the target response magnitude, in the full model. Compare to Figure \ref{fig4}e.
\textbf{f} Median contrast (dev - msc) in the target response magnitude across networks and stimuli in the full and ablated models, arranged as in panel b. The subtraction of the two quantities in the box plot below shows that the target response tends to be increased more in the full model

@@

@@

We had good reasons for this assumption: A priori, updates to the spike-triggered depression variable $x_j$ (cf. Equation \ref{eqn-xsyn}) would decay to a negligible $e^\frac{-1 s}{\tau_x} = 0.0013$ times the original value within two trials, making it unlikely that inputs separated by intervening trials could be "remembered" in the network. Conversely, the frequently presented stimulus B may have been able to lead to a build-up of depression strong enough to suppress its own response.

@@

TODO

@@

\begin{appendices}

\section{Supplementary infomation}\label{secA1}

\begin{figure*}%
    \centering
    \includegraphics{fig4-withSTD.pdf}
    \caption{}
    \label{sfig4}
\end{figure*}

\begin{figure*}%
    \centering
    \includegraphics{fig5-withSTD.pdf}
    \caption{}
    \label{sfig5}
\end{figure*}

\begin{figure*}%
    \centering
    \includegraphics{fig6-withSTD.pdf}
    \caption{}
    \label{sfig6}
\end{figure*}

\begin{figure*}%
    \centering
    \includegraphics{fig8-noTA.pdf}
    \caption{}
    \label{sfig8}
\end{figure*}

@@

\end{appendices}
